---
title: "Feature selection"
author: "Milan Mikuláš, Marek Vatrt"
date: "`r Sys.Date()`"
output: html_document
---

# Goal
In this section we will perform feature selection on our processed data using 3 different techniques.
1. lasso regression
2. elastic net
3. mixed selection

We will then fit the selected features into our best performing model which is logistic regression model. The result will be the best feature selection approach by comparing the model performance.

# Packages

```{r}
packages <- c("glmnet", "pROC", "caret")

install_if_missing <- function(pkgs) {
  missing <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
  if (length(missing) > 0) {
    cat("Installing:", paste(missing, collapse = ", "), "\n")
    install.packages(missing)
  }
}

install_if_missing(packages)
```




# Load data

```{r}
df <- read.csv2("processed.csv")
df <- df[, -1]  # drops the first column (index)

# check if data load was success
dim(df)
```


# Split

Splitting data to train, test to train the Logistic regression model and compare performance of feature selection approaches on test data.

```{r}
library(caret) #createDataPartition
set.seed(67) # replicability
index <- createDataPartition(df$target, p = 0.8, list = FALSE)
train_data <- df[index, ]
test_data  <- df[-index, ]
```

# Help functions

We prepare functions that print out readable summary of model and prepares datasets with features for each feature selection approach.

```{r}

get_selected_features <- function(method_fit){
  coefs <- coef(method_fit, s = "lambda.min")
  selected <- rownames(coefs)[which(coefs != 0)]
  selected <- selected[selected != "(Intercept)"]
  selected
}

get_metrics <- function(tab){
  data.frame(acc=sum(diag(tab)) / sum(tab),
              prec=tab["1", "1"] / sum(tab["1", ]),
              recall=tab["1", "1"] / sum(tab[, "1"]),
              spec=tab["0", "0"] / sum(tab[, "0"]))
}

print_metrics <- function(tab, split_type){
  cat(split_type, " Accuracy:", sum(diag(tab)) / sum(tab), "\n")
  cat(split_type, " Precision:", tab["1", "1"] / sum(tab["1", ]), "\n")
  cat(split_type, " Recall:", tab["1", "1"] / sum(tab[, "1"]), "\n")
  cat(split_type, " Specificity:", tab["0", "0"] / sum(tab[, "0"]), "\n \n")
}

library(pROC) # roc curve

print_model_stats <- function(model, train, test, treshold, name){
  train_prob <- predict(model, type = "response")
  train_pred <- ifelse(train_prob > treshold, 1, 0)
  train_tab <- table(Predicted = train_pred, Actual = train$target)
  
  test_prob <- predict(model, newdata = test, type = "response")
  test_pred <- ifelse(test_prob > treshold, 1, 0)
  test_tab <- table(Predicted = test_pred, Actual = test$target)
 
  # metrics
  metrics <- get_metrics(test_tab)
  #print_metrics(train_tab, "Train")
  #print_metrics(test_tab, "Test")
  
  # confusion matrix
  test_tab_df <- as.data.frame(test_tab)
  
  p <- ggplot(test_tab_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = Freq), size = 8, color = "white") +
    scale_fill_gradient(low = "lightblue", high = "darkblue") +
    theme_minimal() +
    labs(title = name)
  
  print(p)
  
  # roc curve
  roc_obj <- roc(test$target, test_prob)
  plot(roc_obj, main = name,
       col = "blue", lwd = 2,
       print.auc = TRUE, auc.polygon = TRUE)
  
  
  # confidence intervals
  model_summary <- summary(model)$coefficients
  coef_df <- data.frame(
    feature = rownames(model_summary)[-1],
    estimate = model_summary[-1, 1],
    std_error = model_summary[-1, 2],
    p_value = model_summary[-1, 4]
  )
  
  coef_df$significant <- ifelse(coef_df$p_value < 0.05, "Yes", "No")
  coef_df <- coef_df[order(coef_df$estimate), ]
  coef_df$feature <- factor(coef_df$feature, levels = coef_df$feature)
  
  p1 <- ggplot(coef_df, aes(x = estimate, y = feature, color = significant)) +
    geom_point(size = 3) +
    geom_errorbarh(aes(xmin = estimate - 1.96 * std_error,
                       xmax = estimate + 1.96 * std_error), height = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("No" = "grey", "Yes" = "blue")) +
    theme_minimal() +
    labs(title = name,
         x = "Coefficient Estimate", y = "")
  print(p1)
  metrics
}
```


# Lasso

It is a regularization technique that adds an L1 penalty to the logistic regression cost function. The strength of the penalty is controlled by the lambda parameter. Higher lambda values result in more aggressive shrinkage and fewer features, while lower values retain more features at the cost of potential overfitting.


```{r}
library(glmnet) # lasso function

# Prepare data as matrix (glmnet requires this)
X <- as.matrix(train_data[, -which(names(train_data) == "target")])
y <- train_data$target

cv_fit <- cv.glmnet(X, y, family = "binomial", alpha = 1)

selected <- get_selected_features(cv_fit)
lasso_train <- train_data[,c("target", selected)]
lasso_test <- test_data[,c("target", selected)]

```



# Elastic net
Elastic Net combines both L1 (Lasso) and L2 (Ridge) penalties into a single cost function. This is the main advantage over pure Lasso, when features are highly correlated, Lasso tends to pick one and drop the rest randomly. Elastic Net handles this more gracefully by keeping groups of correlated features together while still eliminating truly irrelevant ones. For our experiment we will choose 3 different alphas. Even though our data are not multicolinear, lower alpha can have problem with the model not converging.


```{r}

enet_80 <- cv.glmnet(X, y, family = "binomial", alpha = 0.8)
enet_75 <- cv.glmnet(X, y, family = "binomial", alpha = 0.75)
enet_60 <- cv.glmnet(X, y, family = "binomial", alpha = 0.6)
enet_50 <- cv.glmnet(X, y, family = "binomial", alpha = 0.5)
enet_25 <- cv.glmnet(X, y, family = "binomial", alpha = 0.25)

```




```{r}
compare <- data.frame(
  model = c("Elastic Net 0.8", "Elastic Net 0.75", "Elastic Net 0.6", "Elastic Net 0.5", "Elastic Net 0.25"),
  min_cv_error = c(
    min(enet_80$cvm),
    min(enet_75$cvm),
    min(enet_60$cvm),
    min(enet_50$cvm),
    min(enet_25$cvm)
  ),
  features_selected = c(
    sum(coef(enet_80, s = "lambda.min") != 0) - 1,
    sum(coef(enet_75, s = "lambda.min") != 0) - 1,
    sum(coef(enet_60, s = "lambda.min") != 0) - 1,
    sum(coef(enet_50, s = "lambda.min") != 0) - 1,
    sum(coef(enet_25, s = "lambda.min") != 0) - 1
  )
)

print(compare)
```

```{r}

selected <- get_selected_features(enet_80)
enet_80_train <- train_data[,c("target", selected)]
enet_80_test <- test_data[,c("target", selected)]

selected <- get_selected_features(enet_75)
enet_75_train <- train_data[,c("target", selected)]
enet_75_test <- test_data[,c("target", selected)]

selected <- get_selected_features(enet_60)
enet_60_train <- train_data[,c("target", selected)]
enet_60_test <- test_data[,c("target", selected)]

selected <- get_selected_features(enet_50)
enet_50_train <- train_data[,c("target", selected)]
enet_50_test <- test_data[,c("target", selected)]

selected <- get_selected_features(enet_25)
enet_25_train <- train_data[,c("target", selected)]
enet_25_test <- test_data[,c("target", selected)]
```



# Mixed selection

Mixed (stepwise) selection combines forward and backward selection into a single iterative process. Unlike Lasso and Elastic Net, it doesn't modify the cost function, instead it uses AIC to compare models at each step. Lower AIC means a better balance between fit and complexity.


```{r}
null_model <- glm(target ~ 1, data = train_data, family = "binomial")
full_model <- glm(target ~ ., data = train_data, family = "binomial")

step_model <- step(null_model, 
                   scope = list(lower = null_model, upper = full_model),
                   direction = "both",
                   trace = 1)

selected_features <- names(coef(step_model))[-1]
step_train <- train_data[, c("target", selected_features)]
step_test <- test_data[, c("target", selected_features)]
```

# Comparison


```{r}

datasets <- list(
  lasso   = list(train = lasso_train,   test = lasso_test),
  enet_80 = list(train = enet_80_train, test = enet_80_test),
  enet_75 = list(train = enet_75_train, test = enet_75_test),
  enet_60 = list(train = enet_60_train, test = enet_60_test),
  enet_50 = list(train = enet_50_train, test = enet_50_test),
  enet_25 = list(train = enet_25_train, test = enet_25_test),
  mixed   = list(train = step_train,    test = step_test)
)

models <- lapply(datasets, function(d) {
  glm(target ~ ., data = d$train, family = "binomial")
})

best_acc <- 0
best_model <- NULL
best_name <- NULL

for (name in names(datasets)) {
  d <- datasets[[name]]
  model <- models[[name]]
  metrics <- print_model_stats(model, d$train, d$test, 0.5, name)
  
  if (metrics$acc > best_acc) {
    best_acc <- metrics$acc
    best_model <- model
    best_name <- name
  }
}

cat("Best model:", best_name, "with accuracy:", round(best_acc, 4), "\n")
```

# Conclusion

After using all 3 feature selections methods and comparing them with each other for Logistic regression model it didn't outperform our feature selection done by hand during Logistic regression modeling.



