---
title: "Nonparametric_Models"
author: "Milan Mikuláš, Marek Vatrt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Goal

In this section of our project, our goal is to build 3 non-parametric
models, analyze the results and compare them with each other. To achieve
the best results, we will firstly use default settings for creating the
models and then apply hyperparameter tuning to get the best parameters
(parameters with which the model has the highest accuracy on testing
data).

# Loading libraries and data

```{r, warning=FALSE}
library(tidyverse)
library(tidymodels)

#vif values
library(car)

#Decision tree libs
library(rpart)
library(rpart.plot)
library(caret)

#random forest
library(randomForest)

#k-nn
library(class)

#metrics lib
library(Metrics)
#https://cran.r-project.org/web/packages/Metrics/Metrics.pdf

#clear env variables
rm(list = ls())
```

```{r}
df <- read.csv("processed.csv", sep=";", dec=",")
#removing the first column (exported indexing)
df <- df[2:ncol(df)]

#setting the training margin for train/test split
train_margin=0.8

set.seed(67)
train_index <- sample(1:nrow(df), train_margin * nrow(df))

train <- df[train_index, ]
test  <- df[-train_index, ]

#converting to factor, so the models wont think its for regression
train$target <- as.factor(train$target)
test$target <- as.factor(test$target)
```

Helping function for printing out metrics from our models

```{r}
print_extra_scores <- function(predictions, actual, positive_class = "1") {
  predictions <- factor(predictions, levels = levels(actual))
  actual <- factor(actual, levels = levels(actual))

  if (!(positive_class %in% levels(actual))) {
    positive_class <- levels(actual)[2]
  }
  
  cm <- confusionMatrix(
    data = predictions,
    reference = actual,
    positive = positive_class
  )
  
  cat("\n")
  cat("Accuracy:", round(cm$overall["Accuracy"] * 100, 2), "%\n")
  cat("Precision:", round(cm$byClass["Precision"] * 100, 2), "%\n")
  cat("Recall:", round(cm$byClass["Recall"] * 100, 2), "%\n")
  cat("Specificity:", round(cm$byClass["Specificity"] * 100, 2), "%\n")
  cat("\n\n")
}
```

#Decision tree First non-parametric model we chose is the decision tree.
This choice was made because of the tree's simplicity and
interpretability (although our dataset contains \~55 features, so the
interpretability might be questionable, depending on how the tree will
be created). Since decision trees can be used for regression as well as
classification tasks, we can use it for our task of classifying win/loss
of a team.

## Statistical assumptions

Statistical assumptions are taken from the 5th module. \### Independence
This dataset consists of random matches from the game League of Legends.
Each of the games is independent of eachother, this is because each game
is new. There are no rematch mechanics. \### Multicolinearity Decision
trees can handle multicolinearity, however, features with high
multicollinearity (tested using VIF) were removed from the dataset - as
per advice we got from a consultation. \### Outliers Robust to outliers.
At frist, we did not plan on removing outliers due to the tree being
robust to them, however upon consulting with our supervisor, we have
decided to remove them in preprocessing. \### Sample size May overfit on
small datasets (that is what we want to test) \### Normality, Feature
scaling, Homoscedasticity, Linear relationship Not required.

###Conclusion to statistical assumptions Our dataset satisfies all the
data assumptions of a decision tree. We can now move on to the
implementation of the Decision Tree model.

## Implementation

Sources:
<https://www.rdocumentation.org/packages/rpart/versions/4.1.24/topics/rpart>
<https://www.rdocumentation.org/packages/rpart.plot/versions/3.1.3/topics/rpart.plot>
<https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict>

For the implementation of the decision tree, we will be using the
function rpart(). In this section, we will simply use the default
parameters of rpart. We want to see how the default model compares to
the hyperparameter-tuned model. Also, since decision trees can use Gini
impurity and Entropy scores as the splitting criteria, we will also
compare the accuracy of trees built with Gini and Entropy.

```{r}
#using lib rpart, we create trees for testing the gini inpurity and entropy
decision_tree_model_gini <- rpart(
  target ~ ., 
  data = train,
  #method of classfication
  method = "class",
  #parms is a list of set parameters, even if we just use 1 parameter, rpart expects a list()
  parms=list(split="gini"),
)

decision_tree_model_entropy <- rpart(
  target ~ ., 
  data = train,
  method = "class",
  parms=list(split="information"),
)

#plotting the trees we created
rpart.plot(decision_tree_model_gini,
           main="Gini Inpurity")
rpart.plot(decision_tree_model_entropy,
           main="Entropy")

#using predict() function, we use our test data to make predicitons
predictions_gini <- predict(decision_tree_model_gini, test, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy, test, type = "class")

#accuracy scores on test dataset for gini and entropy
cat("Accuracy on testing dataset (Gini)")
print_extra_scores(predictions_gini, test$target)
cat("Accuracy on testing dataset (Entropy)")
print_extra_scores(predictions_entropy, test$target)

#For final evaluation
base_dt_accuracy <- mean(predictions_gini == test$target)

#predictions on train data - overfitting check
predictions_gini <- predict(decision_tree_model_gini, train, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy, train, type = "class")

#accuracy on training dataset for gini and entropy
cat("Accuracy on training dataset (Gini)")
print_extra_scores(predictions_gini, train$target)
cat("Accuracy on training dataset (Entropy)")
print_extra_scores(predictions_entropy, train$target)

```

As we can see, the decision tree, without even tuning its
hyperparameters, got scores of 95% for accuracy. The more accurate out
of the Gini Inpurity and Entropy methods is the Gini Impurity. For that
reason, we will be using Gini Impurity scoring for tuning the
hyperparameters.

Another fact to mention is that the decision tree is not overfitted.
Actually, for gini impurity, the accuracy score falls behind on the
training dataset by about 0.3%. On the contrary, Entropy score rises on
the training dataset. Both of these scores, however, are very similar to
those of the testing dataset scores. From this we can conclude that the
decision trees are not overfitted.

Since we have decided on the scoring metric we will be using, we can now
move on to tuning the decision tree's hyperparameters. The rpart()
function for creating the decision trees does not have built in
hyperparameters unfortunately. However, we can use rpart.control() to
modify the hyperparameters. The creation and training of decision trees
is not as computanionaly expensive as creating a random forest and
tuning it's hyperparameters, so for the DT, we have decided to use grid
search to search for the best hyperparameters.

The probability of finding a better decision tree is quite slim however.
This is due to the fact that the basic rpart() function has already
found a tree that has a depth of only 3.

Trying to fit trees to a dataset without the most powerful features. We
got these features after manually fitting the rpart() function and
removing features one at a time.

```{r}
#diff_champion_experience is a dominating feature, which splits and finished the decision tree at depth=2
#the accuracy is very high even with only 1 feature deciding the outcome, however, for the sake of exploring relationships between features, we have decided to remove it and try again
exp_cut_train <- train %>%
  select(-c(diff_champion_experience))
exp_cut_test <- test %>%
  select(-c(diff_champion_experience))
```

```{r}
decision_tree_model_gini_exp_cut <- rpart(
  target ~ ., 
  data = exp_cut_train,
  method = "class",
  parms=list(split="gini"),
)

decision_tree_model_entropy_exp_cut <- rpart(
  target ~ ., 
  data = exp_cut_train,
  method = "class",
  parms=list(split="information"),
)

rpart.plot(decision_tree_model_gini_exp_cut,
           main="Gini Inpurity - cut champ exp")
rpart.plot(decision_tree_model_entropy_exp_cut,
           main="Entropy - cut champ exp")

predictions_gini <- predict(decision_tree_model_gini_exp_cut, exp_cut_test, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy_exp_cut, exp_cut_test, type = "class")

#accuracy scores on test dataset for gini and entropy
cat("Accuracy on testing dataset (Gini)")
print_extra_scores(predictions_gini, exp_cut_test$target)
cat("Accuracy on testing dataset (Entropy)")
print_extra_scores(predictions_entropy, exp_cut_test$target)

predictions_gini <- predict(decision_tree_model_gini_exp_cut, exp_cut_train, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy_exp_cut, exp_cut_train, type = "class")

cat("Accuracy on training dataset (Gini)")
print_extra_scores(predictions_gini, exp_cut_train$target)
cat("Accuracy on training dataset (Entropy)")
print_extra_scores(predictions_entropy, exp_cut_train$target)
```

As we can see, removing the most influential feature from the dataset
caused the decision tree to lose a lot of accuracy. What we can also
notice is the fact that the models are starting to overfit more on the
data. There is a more significant difference between testing and
training accuracies (\~4% for Gini and \~5.5% for Entropy).

Since the original DT already has an accuracy score of 97% and depth of
only 2, it is highly unlike that we will find a better DT from
hyperparameter tuning. Still, it is worth to try all combinations, maybe
we will find a more suitable tree that will have more features.

## Hyperparameter tuning

```{r}

dt_iterations <- 50

dt_grid <- data.frame(
  cp = sample(
    c(0.00001, 0.00005, 0.0001, 0.0005, 0.001, 0.005, 0.01, 0.02, 0.05, 0.1),
    dt_iterations,
    replace = TRUE
  ),
  maxdepth = sample(2:30, dt_iterations, replace = TRUE),
  minsplit = sample(2:60, dt_iterations, replace = TRUE)
)

dt_grid$minbucket <- sapply(
  dt_grid$minsplit,
  function(x) sample(1:max(1, floor(x / 2)), 1)
)

dt_grid <- unique(dt_grid)

best_xerror <- Inf
best_params <- NULL
best_cp_for_pruning <- NULL
best_dt_model <- NULL
results <- data.frame()

for (i in 1:nrow(dt_grid)) {

  dt_model <- rpart(
    target ~ .,
    data = train,
    method = "class",
    parms = list(split = "gini"),
    control = rpart.control(
      cp = dt_grid$cp[i],
      maxdepth = dt_grid$maxdepth[i],
      minsplit = dt_grid$minsplit[i],
      minbucket = dt_grid$minbucket[i],
      xval = 5
    )
  )

  model_cptable <- dt_model$cptable

  best_row <- which.min(model_cptable[, "xerror"])

  model_best_xerror <- model_cptable[best_row, "xerror"]
  model_best_cp <- model_cptable[best_row, "CP"]

  results <- rbind(
    results,
    data.frame(
      cp = dt_grid$cp[i],
      maxdepth = dt_grid$maxdepth[i],
      minsplit = dt_grid$minsplit[i],
      minbucket = dt_grid$minbucket[i],
      best_pruning_cp = model_best_cp,
      xerror = model_best_xerror
    )
  )

  if (model_best_xerror < best_xerror) {
    best_xerror <- model_best_xerror
    best_params <- dt_grid[i, ]
    best_cp_for_pruning <- model_best_cp
    best_dt_model <- dt_model
  }
}

results <- results[order(results$xerror, decreasing = FALSE), ]

pruned_best_dt_model <- rpart::prune(
  best_dt_model,
  cp = best_cp_for_pruning
)

test_predictions <- predict(
  pruned_best_dt_model,
  newdata = test,
  type = "class"
)

train_predictions <- predict(
  pruned_best_dt_model,
  newdata = train,
  type = "class"
)

test_accuracy <- mean(test_predictions == test$target)
train_accuracy <- mean(train_predictions == train$target)

#for final eval
tuned_dt_accuracy <- test_accuracy

print(best_params)

cat("Number of random combinations tested:", nrow(dt_grid), "\n")
cat("Best internal CV xerror:", round(best_xerror, 4), "\n")
cat("Best original cp:", best_params$cp, "\n")
cat("Best pruning cp:", best_cp_for_pruning, "\n")
cat("Best maxdepth:", best_params$maxdepth, "\n")
cat("Best minsplit:", best_params$minsplit, "\n")
cat("Best minbucket:", best_params$minbucket, "\n\n")
cat("Training metrics:")
print_extra_scores(train_predictions, train$target)
cat("Testing metrics:")
print_extra_scores(test_predictions, test$target)
```

```{r}
rpart.plot(best_dt_model,
           main="Gini Inpurity - best hyperparameters model")
rpart.plot(decision_tree_model_gini,
           main="Gini Inpurity - default model")
```

As we have suspected, the best model, even with tweaking the
hyperparameters, is still the one created using default rpart()
configuration. The model with tuned hyperparameters achieved better
performance on the training dataset, however, the accuracy score for the
testing dataset remains the same.

```{r}
results %>%
  group_by(maxdepth) %>%
  summarise(
    best_xerror = min(xerror),
    mean_xerror = mean(xerror),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = maxdepth, y = best_xerror)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Decision Tree Tuning: Best CV Error by Tree Depth",
    x = "Max Depth",
    y = "Best CV Error"
  )
```

# Random forest

Building upon the tree models, due to the unexpectedly high accuracy
score of decision tree from both the test and train datasets, we have
decided to choose random forest to see wether we could increase the
accuracy further.

## Statistical assumptions

Statistical assumptions are taken from the 5th module. \### Independence
This dataset consists of random matches from the game League of Legends.
Each of the games is independent of eachother, this is because each game
is new. There are no rematch mechanics. \### Multicolinearity Although
random forest can handle multicolinearity, it is stated that colinearity
could affect feature importance. Due to this reason, we have decided to
get rid of features that have too big of multicolinarity between each
other.

Since we have already cut the features with above 0.8 value on the
correlation matrix during EDA and preprocessing, this time we will focus
on the individual VIF values of the features. We will continue to remove
the feature with the biggest VIF value until the maximum VIF value is
below 10.

Sources:
<https://www.rdocumentation.org/packages/regclass/versions/1.6/topics/VIF>

We ended up only removing 2 features, however, the first feature we
removed had a VIF score of over 4 billion, which is an astronomically
high multicolinearity score. The second feature was over 100 million.
From the scores alone, we can asses that it was a good decision to
remove them from the dataset before creating a random forest model.

### Outliers

Robust to outliers \### Sample size Performs better with a larger
dataset. Our dataset is not very large (\~1200 rows), however, due to
the accuracy of the decision tree we tested, we believe that the random
forest will still perform well. \### Normality, Feature scaling,
Homoscedasticity, Linear relationship Not required

## Implementation

### No tuning, default

```{r}
random_forest_model <- randomForest(
  target ~ .,
  data = train,
  replace = TRUE,
  importance = TRUE
)

random_forest_model

rf_predictions <- predict(
  random_forest_model,
  newdata = test,
  type = "class"
)

table(
  Predicted = rf_predictions,
  Actual = test$target
)

rf_accuracy <- mean(rf_predictions == test$target)

cat("Random Forest testing metrics")
print_extra_scores(rf_predictions, test$target)
```

As we can see. Just by using the default values (+ setting the
replacement and importance to TRUE), we got up to 95.34% accuracy. This
score is outstanding, but we are wondering if we can do even better by
searching through multiple combinations of hyperparameters to find a
better result. We have decided to use random search for hyperparameters.
We could also use grid search, however, finding the best combinations of
hyperparameters with grid search could become very exhaustive,
especially with the \~55 features we have. Random search could find a
better combination of hyperparameters much faster because it does not
waste time trying every combination that exists.

At first, we tried using the train() function, however, the only
hyperparameter we are able to tweak is the **mtry** parameter. Mtry is
important, but this kind of random testing only provided minor
improvements (around 0.5% accuracy). We are wondering if it is possible
to go beyond that. In this instance, we will not be using a built-in
function, instead we will try to do our own version of a randomized
search, where we try to tweak all the numeric parameters of the random
forest - mtree, mtry, nodesize, maxnodes using the randomForest library
and cross validation.

## Hyperparameter tuning

```{r}
tune_random_forest <- function(
  train,
  test,
  target_col = "target",
  seed = 67,
  rf_iterations = 50
) {

  set.seed(seed)

  predictor_count <- ncol(train) - 1

  random_grid <- data.frame(
    mtry = sample(1:predictor_count, rf_iterations, replace = TRUE),
    ntree = sample(seq(100, 1000, by = 50), rf_iterations, replace = TRUE),
    nodesize = sample(20:100, rf_iterations, replace = TRUE),
    maxnodes = sample(seq(10, 60, by = 2), rf_iterations, replace = TRUE)
  )

  random_grid <- unique(random_grid)

  best_oob_accuracy <- 0
  best_params <- NULL
  results <- data.frame()

  for (i in 1:nrow(random_grid)) {

    rf_model <- randomForest(
      as.formula(paste(target_col, "~ .")),
      data = train,
      mtry = random_grid$mtry[i],
      ntree = random_grid$ntree[i],
      nodesize = random_grid$nodesize[i],
      maxnodes = random_grid$maxnodes[i],
      replace = TRUE,
      importance = TRUE
    )

    final_oob_error <- rf_model$err.rate[rf_model$ntree, "OOB"]
    oob_accuracy <- 1 - final_oob_error

    results <- rbind(
      results,
      data.frame(
        mtry = random_grid$mtry[i],
        ntree = random_grid$ntree[i],
        nodesize = random_grid$nodesize[i],
        maxnodes = random_grid$maxnodes[i],
        oob_accuracy = oob_accuracy,
        oob_error = final_oob_error
      )
    )

    if (oob_accuracy > best_oob_accuracy) {
      best_oob_accuracy <- oob_accuracy
      best_params <- random_grid[i, ]
    }
  }

  results <- results[order(results$oob_accuracy, decreasing = TRUE), ]

  set.seed(seed)

  final_model <- randomForest(
    as.formula(paste(target_col, "~ .")),
    data = train,
    mtry = best_params$mtry,
    ntree = best_params$ntree,
    nodesize = best_params$nodesize,
    maxnodes = best_params$maxnodes,
    replace = TRUE,
    importance = TRUE
  )

  train_predictions <- predict(final_model, newdata = train)
  test_predictions <- predict(final_model, newdata = test)

  train_accuracy <- mean(train_predictions == train[[target_col]])
  test_accuracy <- mean(test_predictions == test[[target_col]])

  final_oob_error <- final_model$err.rate[final_model$ntree, "OOB"]
  final_oob_accuracy <- 1 - final_oob_error

  cat("Best tuning OOB accuracy:", round(best_oob_accuracy * 100, 2), "%\n")
  cat("Final model OOB accuracy:", round(final_oob_accuracy * 100, 2), "%\n\n")
  cat("Training metrics:")
  print_extra_scores(train_predictions, train[[target_col]])
  cat("Testing metrics:")
  print_extra_scores(test_predictions, test[[target_col]])
  cat("Train-test gap:", round((train_accuracy - test_accuracy) * 100, 2), "%\n")
  cat("Best mtry:", best_params$mtry, "\n")
  cat("Best ntree:", best_params$ntree, "\n")
  cat("Best nodesize:", best_params$nodesize, "\n")
  cat("Best maxnodes:", best_params$maxnodes, "\n")

  return(list(
    final_model = final_model,
    best_params = best_params,
    best_tuning_oob_accuracy = best_oob_accuracy,
    final_oob_accuracy = final_oob_accuracy,
    train_accuracy = train_accuracy,
    test_accuracy = test_accuracy,
    results = results
  ))
}
```

The best outcome we got from this hyperparameter tuning after we tried
changing all the randomised ranges of the hyperparameters was 96.99%.
That is an upgrade of 1.1% compared to the train() function. What we
have not done yet is checking for overfitting. We will now test the best
model on both the train dataset and the test dataset and compare their
accuracy.

```{r}
rf_tuned <- tune_random_forest(train, test)
```

```{r}
plot_rf_param <- function(parameter) {
  ggplot(rf_tuned$results, aes(x = .data[[parameter]], y = oob_accuracy)) +
    geom_point(alpha = 0.5) +
    geom_smooth(se = FALSE) +
    labs(
      title = paste("Random Forest Tuning:", parameter, "vs OOB Accuracy"),
      x = parameter,
      y = "OOB Accuracy"
    ) +
    scale_y_continuous(labels = scales::percent)
}

```

```{r}
plot_rf_param("mtry")
plot_rf_param("ntree")
plot_rf_param("nodesize")
plot_rf_param("maxnodes")
```

## Hyperparameter tuning conclusion

The most succesful model came from our own implementation of random
search with an accuracy of 96.99%. Even though it might not seem like a
substanstial difference, the "baseline" for this model was already very
high at 95.34%.

### Feature importance

Now that we have found the best model via hyperparameter tuning, we will
look into what are the most important, or valuable features for
splitting the trees.

Sources:
<https://www.rdocumentation.org/packages/randomForest/versions/4.7-1.2/topics/varImpPlot>

```{r, fig.width=16, fig.height=8}
varImpPlot(
  rf_tuned$final_model,
  n.var = 10,
  main = "Hyperparameter-tuned RF Feature Importance"
)
```

As we can see, the same feature that dominated the decision tree
dominated again in the random forest.

diff_inhibitor_takedowns, without being a data leakage, is the most
significant and influential variable for predicting the outcome of the
match. It has the highest MeanDecreaseAccuracy, meaning that randomly
permuting this feature causes the largest drop in classification
accuracy. It also has the highest MeanDecreseGini, which means that it
contributes the most to the reduction of gini impurity at a split.

We debated wether to remove it, simply because it carries a lot of
influencer over the outcome. However, as the decision tree has proved,
even if we were to remove it, other features can still classify the
outcome of the matches with a great accuracy. In the end, we have
decided to keep it, because it is a part of the dataset and it is not
considered a data leakage.

#k-NN As the final non-parametric model we have chosen k-NN. At first we
wanted to use XGBoost, because we thought that the decision tree and
random forest would not be as powerful, so we wanted to use boosting to
combine the "weaker" trees into a better one. However, the accuracy of
the tree models turned out to be excellent, that is why we decided to
move on from tree algorithms.

k-NN is simple, and unlike tree algorithms, that split the feature space
based on conditions, k-NN uses the similarity of data to make decisions.
We wanted to choose it due to the use of a different technique. However,
k-NN has a defining problem, sometimes called the "curse of
dimensionality". What this means is that if a dataset contains too many
features, the distances between individual features can become similar,
which can cause the model to predict unreliably. This is a big problem
for our dataset, which contains \~55 features.

To counter this problem, we are planning on implementing a simple
feature selection, to select the best 5-15 features. We plan to do this
using the most random forest's feature importance. Using the top 15
features, we will then observe how the accuracy of k-NN
increases/decreases when we try using 14,13,12,..,5 features.

What we found from using the tree models is the fact that our dataset
contains a small set of features, which carry the most influence over
the outcome of the prediction. We want to see exactly how much the
number of features changes the accuracy and if a different algorithm can
bring even more accuracy to our predictions.

## Statistical assumptions

Sources for statistical assumptions of k-NN: \n
<https://medium.com/analytics-vidhya/k-nearest-neighbors-algorithm-7952234c69a4>
\n

### Independence

This dataset consists of random matches from the game League of Legends.
Each of the games is independent of eachother, this is because each game
is new. There are no rematch mechanics. \### Multicollinearity
Multicollinearity can affect k-NN performance, we can try eliminating
the features with high multicollinearity, just like we did in RF. \###
Outliers Sensitive to outliers. \### Sample size Computationally
expensive on large datasets, could overfit on small dataset. \###
Normality, Homoscedasticity, Linear relationship Not required \###
Feature scaling Due to k-NN being a distance-based method of
classification, feature scaling is required.

```{r}
#k-nn
library(class)
library(kknn)
```

##Implementation Base model with default hyperparameters

```{r}
library(kknn)
model <- kknn(
          target ~ .,
          train,
          test
        )

pred <- predict(model)
cat("Base k-NN metrics: ")
print_extra_scores(pred, test$target)
```

##Hyperparameter tuning The hyperparameters we will be tuning to find
the best k-NN model is the k value, meaning the number of neighbours
needed for classifying a point, distance, and kernel. We can choose
distances 1,2,3 - Manhattan, Euclidian and Minkowski. There are 10
kernels available for kknn, we chose to try all of them. The final
number of combinations adds up to 1500.

We will iterate through all the combinations - grid search, since there
is "only" 1500 combinations, doing this will not be as computationaly
expensive as random forest, for which we opted to use random search.

### Hyperparameter search

```{r}
train_copy <- train
test_copy <- test
```

```{r}
tune_knn <- function(
  train,
  test,
  target_col = "target",
  seed = 67,
  knn_iterations = 50,
  #when we lowered the k-values, the model found good solutions, but trainig set accuracy score was 100% - overfit
  k_values = 1:50,
  distance_metrics = 1:3,
  kernels = c(
    "rectangular",
    "triangular",
    "epanechnikov",
    "biweight",
    "triweight",
    "cos",
    "inv",
    "gaussian",
    "rank",
    "optimal"),
  k_folds = 5
) {

  set.seed(seed)

  random_grid <- data.frame(
    k = sample(k_values, knn_iterations, replace = TRUE),
    distance = sample(distance_metrics, knn_iterations, replace = TRUE),
    kernel = sample(kernels, knn_iterations, replace = TRUE)
  )

  #keeping only the unique random combinations=no duplicates
  random_grid <- unique(random_grid)

  folds <- createFolds(
    train[[target_col]],
    k = k_folds,
    list = TRUE,
    returnTrain = FALSE
  )

  best_k <- NULL
  best_distance <- NULL
  best_kernel <- NULL
  best_cv_accuracy <- 0
  results <- data.frame()

  for (i in 1:nrow(random_grid)) {

    fold_accuracies <- numeric(k_folds)

    for (fold_id in 1:k_folds) {

      validation_index <- folds[[fold_id]]

      cv_train_data <- train[-validation_index, ]
      cv_validation_data <- train[validation_index, ]

      cv_train_data[[target_col]] <- as.factor(cv_train_data[[target_col]])
      cv_validation_data[[target_col]] <- as.factor(cv_validation_data[[target_col]])

      model <- kknn(
        target ~ .,
        train = cv_train_data,
        test = cv_validation_data,
        k = random_grid$k[i],
        distance = random_grid$distance[i],
        kernel = random_grid$kernel[i],
        #when we did not use scaling, the accuracy was the same for all combinations
        scale = TRUE
      )

      predictions <- predict(model)

      fold_accuracies[fold_id] <- mean(
        predictions == cv_validation_data[[target_col]]
      )
    }

    cv_accuracy <- mean(fold_accuracies)

    results <- rbind(
      results,
      data.frame(
        k = random_grid$k[i],
        kernel = random_grid$kernel[i],
        distance = random_grid$distance[i],
        cv_accuracy = cv_accuracy
      )
    )

    if (cv_accuracy > best_cv_accuracy) {
      best_cv_accuracy <- cv_accuracy
      best_k <- random_grid$k[i]
      best_distance <- random_grid$distance[i]
      best_kernel <- random_grid$kernel[i]
    }
  }

  results <- results[order(results$cv_accuracy, decreasing = TRUE), ]

  best_model <- kknn(
    as.formula(paste(target_col, "~ .")),
    train = train,
    test = test,
    k = best_k,
    distance = best_distance,
    kernel = best_kernel,
    scale = TRUE
  )

  test_predictions <- predict(best_model)
  test_accuracy <- mean(test_predictions == test[[target_col]])

  generalization_gap <- best_cv_accuracy - test_accuracy

  cat("Number of random combinations tested:", nrow(random_grid), "\n")
  cat("Best CV accuracy:", round(best_cv_accuracy * 100, 2), "%\n")
  cat("Testing metrics:")
  print_extra_scores(test_predictions, test[[target_col]])
  cat("CV-test gap:", round(generalization_gap * 100, 2), "%\n")
  cat("Best k:", best_k, "\n")
  cat("Best distance metric:", best_distance, "\n")
  cat("Best kernel:", best_kernel, "\n")

  return(list(
    best_model = best_model,
    best_k = best_k,
    best_distance = best_distance,
    best_kernel = best_kernel,
    best_cv_accuracy = best_cv_accuracy,
    test_accuracy = test_accuracy,
    generalization_gap = generalization_gap,
    results = results,
    random_grid = random_grid,
    train = train,
    test = test
  ))
}
```

```{r}
knn_tuned_all_features <- tune_knn(train_copy,test_copy)
```

```{r}
ggplot(knn_tuned_all_features$results, aes(x = k, y = cv_accuracy)) +
  geom_point(alpha = 0.6) +
  geom_line() +
  labs(
    title = "k-NN Tuning: Number of Neighbors vs CV Accuracy",
    x = "k",
    y = "Cross-validation Accuracy"
  ) +
  scale_y_continuous(labels = scales::percent)

ggplot(knn_tuned_all_features$results, aes(x = kernel, y = cv_accuracy, group = 1)) +
  geom_point(alpha = 0.6) +
  geom_line() +
  labs(
    title = "k-NN Tuning: Kernel Comparison",
    x = "Kernel",
    y = "CV Accuracy"
  ) +
  scale_y_continuous(labels = scales::percent)
```

We also tried to use train() and trainControl() to control the cross
validation and training of the models. Due to the different creation of
folds for the training and validation of the models, the outcomes
slightly differed. We achieved similar results in terms of accuracy,
however, the training time of all the models took 5x longer using the
train() function (2 hours), so we have decided to choose our
implementation.

## PCA

Since our dataset contains \~55 features, we have decided to try using a
dimensionality reduction method to lower the number of features. For
this, we have chosen PCA.

```{r}
tune_knn_pca <- function(
  train,
  test,
  target_col = "target",
  seed = 67,
  pca_iterations = 50,
  pca_thresholds = seq(0.39, 0.99, by = 0.05),
  k_values = 10:50,
  distance_metrics = 1:3,
  kernels = c(
    "rectangular",
    "triangular",
    "epanechnikov",
    "biweight",
    "triweight",
    "cos",
    "inv",
    "gaussian",
    "rank",
    "optimal"
  ),
  k_folds = 5
) {

  set.seed(seed)

  x_train <- train[, names(train) != target_col]
  y_train <- train[[target_col]]

  x_test <- test[, names(test) != target_col]
  y_test <- test[[target_col]]

  random_grid <- data.frame(
    pca_threshold = sample(pca_thresholds, pca_iterations, replace = TRUE),
    k = sample(k_values, pca_iterations, replace = TRUE),
    distance = sample(distance_metrics, pca_iterations, replace = TRUE),
    kernel = sample(kernels, pca_iterations, replace = TRUE)
  )

  random_grid <- unique(random_grid)

  folds <- createFolds(
    y_train,
    k = k_folds,
    list = TRUE,
    returnTrain = FALSE
  )

  best_cv_accuracy <- 0
  best_params <- NULL
  results <- data.frame()

  for (i in 1:nrow(random_grid)) {

    fold_accuracies <- numeric(k_folds)
    fold_components <- numeric(k_folds)

    for (fold_id in 1:k_folds) {

      validation_index <- folds[[fold_id]]

      cv_x_train <- x_train[-validation_index, ]
      cv_y_train <- y_train[-validation_index]

      cv_x_validation <- x_train[validation_index, ]
      cv_y_validation <- y_train[validation_index]

      pca_model <- preProcess(
        cv_x_train,
        method = c("center", "scale", "pca"),
        thresh = random_grid$pca_threshold[i]
      )

      cv_x_train_pca <- predict(pca_model, cv_x_train)
      cv_x_validation_pca <- predict(pca_model, cv_x_validation)

      fold_components[fold_id] <- ncol(cv_x_train_pca)

      cv_train_pca <- data.frame(cv_x_train_pca)
      cv_validation_pca <- data.frame(cv_x_validation_pca)

      cv_train_pca[[target_col]] <- cv_y_train
      cv_validation_pca[[target_col]] <- cv_y_validation

      model <- kknn(
        as.formula(paste(target_col, "~ .")),
        train = cv_train_pca,
        test = cv_validation_pca,
        k = random_grid$k[i],
        distance = random_grid$distance[i],
        kernel = random_grid$kernel[i]
      )

      predictions <- predict(model)

      fold_accuracies[fold_id] <- mean(
        predictions == cv_validation_pca[[target_col]]
      )
    }

    cv_accuracy <- mean(fold_accuracies)
    mean_components <- mean(fold_components)

    results <- rbind(
      results,
      data.frame(
        pca_threshold = random_grid$pca_threshold[i],
        k = random_grid$k[i],
        distance = random_grid$distance[i],
        kernel = random_grid$kernel[i],
        cv_accuracy = cv_accuracy,
        mean_components = mean_components
      )
    )

    if (cv_accuracy > best_cv_accuracy) {
      best_cv_accuracy <- cv_accuracy
      best_params <- random_grid[i, ]
    }
  }

  results <- results[order(results$cv_accuracy, decreasing = TRUE), ]

  final_pca_model <- preProcess(
    x_train,
    method = c("center", "scale", "pca"),
    thresh = best_params$pca_threshold
  )

  x_train_pca <- predict(final_pca_model, x_train)
  x_test_pca <- predict(final_pca_model, x_test)

  knn_train_pca <- data.frame(x_train_pca)
  knn_test_pca <- data.frame(x_test_pca)

  knn_train_pca[[target_col]] <- y_train
  knn_test_pca[[target_col]] <- y_test

  final_model <- kknn(
    as.formula(paste(target_col, "~ .")),
    train = knn_train_pca,
    test = knn_test_pca,
    k = best_params$k,
    distance = best_params$distance,
    kernel = best_params$kernel
  )

  test_predictions <- predict(final_model)
  test_accuracy <- mean(test_predictions == knn_test_pca[[target_col]])

  final_components <- ncol(x_train_pca)

  generalization_gap <- best_cv_accuracy - test_accuracy

  cat("Number of random combinations tested:", nrow(random_grid), "\n")
  cat("Best CV accuracy:", round(best_cv_accuracy * 100, 2), "%\n")
  cat("Testing metrics:")
  print_extra_scores(test_predictions, knn_test_pca[[target_col]])
  cat("CV-test gap:", round(generalization_gap * 100, 2), "%\n")
  cat("Best PCA threshold:", best_params$pca_threshold, "\n")
  cat("Number of PCA components:", final_components, "\n")
  cat("Best k:", best_params$k, "\n")
  cat("Best distance:", best_params$distance, "\n")
  cat("Best kernel:", best_params$kernel, "\n")

  return(list(
    final_model = final_model,
    final_pca_model = final_pca_model,
    best_params = best_params,
    best_cv_accuracy = best_cv_accuracy,
    test_accuracy = test_accuracy,
    generalization_gap = generalization_gap,
    final_components = final_components,
    results = results,
    random_grid = random_grid,
    train_pca = knn_train_pca,
    test_pca = knn_test_pca
  ))
}
```

```{r}
knn_pca_tuned <- tune_knn_pca(
  train = train,
  test = test,
  pca_iterations = 50
)
```

```{r}
# pca_results_summary <- knn_pca_tuned$results %>%
#   group_by(pca_threshold) %>%
#   summarise(
#     mean_accuracy = mean(cv_accuracy, na.rm = TRUE),
#     best_accuracy = max(cv_accuracy, na.rm = TRUE),
#     sd_accuracy = sd(cv_accuracy, na.rm = TRUE),
#     mean_components = mean(mean_components, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot(pca_results_summary, aes(x = pca_threshold, y = mean_accuracy)) +
#   geom_line(linewidth = 1) +
#   geom_point(size = 2) +
#   labs(
#     title = "PCA k-NN: PCA Threshold vs Mean CV Accuracy",
#     x = "PCA explained variance threshold",
#     y = "Mean CV Accuracy"
#   ) +
#   scale_y_continuous(labels = scales::percent)
```

```{r}
pca_components_summary <- knn_pca_tuned$results %>%
  group_by(mean_components) %>%
  summarise(
    mean_accuracy = mean(cv_accuracy, na.rm = TRUE),
    best_accuracy = max(cv_accuracy, na.rm = TRUE),
    sd_accuracy = sd(cv_accuracy, na.rm = TRUE),
    tested_combinations = n(),
    .groups = "drop"
  )

ggplot(pca_components_summary, aes(x = mean_components, y = mean_accuracy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_accuracy - sd_accuracy,
      ymax = mean_accuracy + sd_accuracy
    ),
    width = 0.3,
    alpha = 0.5
  ) +
  labs(
    title = "PCA k-NN: Number of PCA Components vs Mean CV Accuracy",
    x = "Mean number of PCA components",
    y = "Mean CV Accuracy"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

```

# Final Evalutation

```{r}
library(pROC)

get_positive_prob <- function(prob_object, positive_class = "1") {
  prob_df <- as.data.frame(prob_object)
  
  if (positive_class %in% colnames(prob_df)) {
    return(prob_df[[positive_class]])
  }
  
  return(prob_df[[ncol(prob_df)]])
}

dt_prob <- predict(
  pruned_best_dt_model,
  newdata = test,
  type = "prob"
)[, "1"]

rf_prob <- predict(
  rf_tuned$final_model,
  newdata = test,
  type = "prob"
)[, "1"]

knn_prob <- get_positive_prob(
  predict(knn_tuned_all_features$best_model, type = "prob"),
  positive_class = "1"
)

knn_pca_prob <- get_positive_prob(
  predict(knn_pca_tuned$final_model, type = "prob"),
  positive_class = "1"
)

roc_dt <- pROC::roc(test$target, dt_prob, levels = c("0", "1"), direction = "<")
roc_rf <- pROC::roc(test$target, rf_prob, levels = c("0", "1"), direction = "<")
roc_knn <- pROC::roc(test$target, knn_prob, levels = c("0", "1"), direction = "<")
roc_knn_pca <- pROC::roc(
  knn_pca_tuned$test_pca$target,
  knn_pca_prob,
  levels = c("0", "1"),
  direction = "<"
)

roc_data <- bind_rows(
  data.frame(
    fpr = 1 - roc_dt$specificities,
    tpr = roc_dt$sensitivities,
    model = paste0("Decision Tree tuned, AUC = ", round(as.numeric(pROC::auc(roc_dt)), 3))
  ),
  data.frame(
    fpr = 1 - roc_rf$specificities,
    tpr = roc_rf$sensitivities,
    model = paste0("Random Forest tuned, AUC = ", round(as.numeric(pROC::auc(roc_rf)), 3))
  ),
  data.frame(
    fpr = 1 - roc_knn$specificities,
    tpr = roc_knn$sensitivities,
    model = paste0("k-NN tuned, AUC = ", round(as.numeric(pROC::auc(roc_knn)), 3))
  ),
  data.frame(
    fpr = 1 - roc_knn_pca$specificities,
    tpr = roc_knn_pca$sensitivities,
    model = paste0("k-NN PCA tuned, AUC = ", round(as.numeric(pROC::auc(roc_knn_pca)), 3))
  )
)

ggplot(roc_data, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 1.1) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "gray40"
  ) +
  coord_equal() +
  labs(
    title = "ROC Curve Comparison of Tuned Models",
    subtitle = "Curves closer to the top-left corner indicate better classification performance",
    x = "False Positive Rate",
    y = "True Positive Rate / Recall",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

cat("AUC - Decision Tree tuned:", round(as.numeric(pROC::auc(roc_dt)), 4), "\n")
cat("AUC - Random Forest tuned:", round(as.numeric(pROC::auc(roc_rf)), 4), "\n")
cat("AUC - k-NN tuned:", round(as.numeric(pROC::auc(roc_knn)), 4), "\n")
cat("AUC - k-NN PCA tuned:", round(as.numeric(pROC::auc(roc_knn_pca)), 4), "\n")
```

```{r}
model_results <- data.frame(
  model = c(
    "Decision Tree default",
    "Decision Tree tuned",
    "Random Forest default",
    "Random Forest tuned",
    "k-NN base model",
    "k-NN all features tuned",
    "k-NN PCA tuned"
  ),
  test_accuracy = as.numeric(c(
    base_dt_accuracy,
    tuned_dt_accuracy,
    rf_accuracy,
    rf_tuned$test_accuracy,
    base_knn_accuracy,
    knn_tuned_all_features$test_accuracy,
    knn_pca_tuned$test_accuracy
  ))
)

ggplot(model_results, aes(x = reorder(model, test_accuracy), y = test_accuracy)) +
  geom_col() +
  coord_flip(ylim = c(0.80, 1.00)) +
  labs(
    title = "Comparison of Final Model Test Accuracies",
    x = "Model",
    y = "Test Accuracy"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    breaks = seq(0.80, 1.00, by = 0.05)
  )
```
