---
title: "Parametric_Models"
author: "Milan Mikuláš, Marek Vatrt"
date: "`r Sys.Date()`"
output: html_document
---

# Goal

Our goal for this notebook is to build 3 different statistical parametric models for predicting win (binary classification) on our League of Legends dataset. We aim to then compare these models based on metrics to find our which model performs the best

# Data Load

```{r}
df <- read.csv2("processed.csv")
df <- df[, -1]  # drops the first column (index)

# check if data load was success
dim(df)
```

# Split

To correctly measure our metrics we need to split the data to train and test datasets. We split our data in 80/20 ratio. We also set seed for replicability

```{r}
library(caret) #createDataPartition
set.seed(67) # replicability
index <- createDataPartition(df$target, p = 0.8, list = FALSE)
train_data <- df[index, ]
test_data  <- df[-index, ]
```

# Models

We choose 3 different parametric models. Each has its own sets of conditions the predictors have to follow. Because of that we need to treat each model preparation as its own data preprocessing and feature selection

We decided to try - Logistic regression - Naive Bayes - LDA

# Logistic regression

## Assumptions of Logicstic Regression

1.  Independent observations -\> each observation is a different Match - they are independent
2.  Binary dependent variables -\> Target is binary (team100 1-win 0-lose -\> team200 won)
3.  Large sample size -\> our train data has \~1000 rows which we consulted and is enough
4.  Linearity relationship between independent variables and log odds
5.  No extreme outliers
6.  Colinearity

We will deal with 4.-6. in preprocessing

## Preprocessing

Each section of data processing potentially losses predictor for now following the model conditions. We progress in these sections in a way, that minimize losing predictors to the most as we can.

### Log-odds relationship

In a non-formal way, this assumption means increasing feature value also increases the probability of it. We can check and see this visually. We plot a smooth curve on a plot where y is our target 0/1 and x has our feature. We select features that copy a sigmoid curve.

```{r}
plot_logistic <- function(df, feature) {
  ggplot(df, aes(x = (.data[[feature]]), y = target)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
    labs(x = feature, y = "P(target = 1)") +
    theme_minimal()
}
```

```{r}
lapply(names(train_data), function(f) plot_logistic(train_data, f))
```

After observing each feature we decided to keep these following features

```{r}
predictors <- c("diff_assists",
                "diff_triple_kills",
                "diff_largest_multi_kill",
                "diff_killing_sprees",
                "diff_total_damage_dealt",
                "diff_total_damage_taken",
                "diff_physical_damage_dealt",
                "diff_true_damage_dealt",
                "diff_gold_spent",
                "diff_total_enemy_jungle_minions_killed",
                "diff_dragon_kills",
                "diff_baron_kills",
                "diff_inhibitor_takedowns",
                "diff_total_heal",
                "diff_solo_kills",
                "diff_outnumbered_kills"
)

train_data_lr <- train_data[, c("target", predictors)]
test_data_lr <- test_data[, c("target", predictors)]
```

### Colinearity

We already dropped any features that had \|correlation\| \>= 0.8 in eda + preprocessing. What we need to do further is to check VIF and iteratively drop features with VIF \> 5.

```{r}
library(car)
remove_vif <- function(df){
  current_predictors <- setdiff(names(df), "target")
  repeat {
    current_formula <- as.formula(
      paste("target ~", paste(current_predictors, collapse = " + "))
    )
    current_model <- glm(current_formula, data = df, family = "binomial")
    
    vif_vals <- vif(current_model)
    max_vif <- max(vif_vals)
    
    if (max_vif <= threshold) break
    
    drop <- names(which.max(vif_vals))
    current_predictors <- setdiff(current_predictors, drop)
  }
  return(current_predictors)
}
```

Before doing VIF check

```{r}
length(names(train_data_lr))
```

After doing VIF check

```{r}
new_pred <- remove_vif(train_data_lr)
length(new_pred)
```

We only had to drop 1 feature

```{r}
train_data_lr <- train_data_lr[, c("target", new_pred)]
test_data_lr <- test_data_lr[, c("target", new_pred)]
```

### Extreme outliers

We decided to declare extreme outliers to be beyond 3\*IQR. We first check the outliers and than decide what to do. There could be some issues we dont want to just drop or clip any values if possible.

```{r}
detect_extreme_outliers <- function(df){
  for (feat in names(df)) {
    x <- df[[feat]]
    Q1 <- quantile(x, 0.25)
    Q3 <- quantile(x, 0.75)
    IQR_val <- Q3 - Q1
      
    extreme <- which(x < Q1 - 3 * IQR_val | x > Q3 + 3 * IQR_val)
      
    if (length(extreme) > 0) {
      cat(sprintf("%-45s extreme outliers: %d (%.2f%%)\n",
                  feat, length(extreme),
                  length(extreme) / length(x) * 100))
    }
  }
}
```

```{r}
detect_extreme_outliers(train_data_lr)
```

```{r}
hist(train_data_lr$diff_triple_kills)
boxplot(train_data_lr$diff_triple_kills)
```

For this feature we dont count these values as extreme outliers. The distribution doesnt look as bad as the boxplot shows. We will keep the feature and clip the other. Cap at 3 standard deviations

```{r}
for (feat in names(train_data_lr)) {
  x <- train_data_lr[[feat]]
  mu <- mean(x)
  s <- sd(x)
  train_data_lr[[feat]] <- pmax(pmin(x, mu + 3*s), mu - 3*s)
  
  # Apply same caps to test set using TRAIN mean/sd
  test_data[[feat]] <- pmax(pmin(test_data[[feat]], mu + 3*s), mu - 3*s)
}
```

## Model fit

```{r}
log_regresion_model <- glm(target ~ ., data = train_data_lr, family = "binomial")
```

## Summary

We can see that out of our 16 predictors only 7 are significant. Test accuracy is 0.9671 and precision 0.9558. Our model doesnt seem to be overfitted either after checking train accuracy being 0.9661

```{r}
accuracy <- function(tab){
  return(sum(diag(tab)) / sum(tab))
}

precision <- function(tab){
  return(tab["1", "1"] / sum(tab["1", ]))
}
```

```{r}
lrm_test_prob <- predict(log_regresion_model, newdata = test_data_lr, type = "response")

lrm_test_pred <- ifelse(lrm_test_prob > 0.5, 1, 0)

test_tab <- table(Predicted = lrm_test_pred, Actual = test_data_lr$target)
print(test_tab)

acc <- accuracy(test_tab)
prec <- precision(test_tab)

cat("Accuracy:", round(acc, 4), "\n")
cat("Precision:", round(prec, 4), "\n")

summary(log_regresion_model)
```

# Naive Bayes

## Assumptions of NB

<https://www.geeksforgeeks.org/machine-learning/naive-bayes-classifiers/>

1.  Feature independence

-   this is a "Naive" assumption that is violated in real world
-   we already dropped highly correlated features
-   so we will just move on

2.  Continuous features are normally distributed

-   we have to check normality for each class

3.  Discrete features have multinomial distributions

-   <https://faculty.washington.edu/yenchic/20A_stat512/Lec7_Multinomial.pdf> 
4. Features are equally important
5. No missing data

## Preprocessing
We need to check normality for continuous data for each class. Lets first split data into discrete and continuous. Lets look at the qqplots of features and decide which are continuous and discrete


```{r}
par(mfrow = c(1, 1))
predictors <- names(train_data)
for (feat in predictors) {
  qqnorm(train_data[[feat]], main = feat, cex.main = 0.8)
  qqline(train_data[[feat]], col = "red")
}
```

After observing the qqplots we decided to split the data 

```{r}
discrete <- c("target", "diff_triple_kills", "diff_largest_multi_kill",
              "diff_dragon_kills", "diff_baron_kills",
              "diff_objectives_stolen", "diff_first_blood_kill",
              "diff_first_blood_assist", "diff_first_tower_kill",
              "diff_first_tower_assist", "diff_wards_guarded")

continuous <- setdiff(names(train_data), discrete)
```


```{r}
cat("\n Discrete \n")
for (feat in discrete) {
  cat(feat, length(unique(train_data[[feat]])), "\n")
}

cat("\n Continuous \n")
for (feat in continuous) {
  cat(feat, length(unique(train_data[[feat]])), "\n")
}
```
We can see that we devided the data into groups where discrete features have < 20 unique values.


### Target class split
For NB the normality is check for each class so we need to split the data. We also just need to check the continuous data so lets only include them.

```{r}
class_0 <- train_data[train_data$target == 0, continuous]
class_1 <- train_data[train_data$target == 1, continuous]
```



### Normality check
We will use both visual techniques and shapiros willk test based on paper https://pmc.ncbi.nlm.nih.gov/articles/PMC3693611/


```{r}
library(e1071)
check_normality <- function(df, features){
  results <- data.frame(
    feature = character(),
    shapiro_p = numeric(),
    skewness = numeric(),
    kurtosis = numeric(),
    qq_verdict = character(),
    final_verdict = character(),
    stringsAsFactors = FALSE
  )
  
  pass <- df[, features]
  fail <- df[, features]
  
  for (feat in features) {
    x <- train_data[[feat]][!is.na(train_data[[feat]])]
    
    sh <- shapiro.test(x)
    sk <- skewness(x)
    kt <- kurtosis(x)
    
    # Practical normality from skewness/kurtosis
    practical <- abs(sk) < 1 & abs(kt) < 3
    if (practical == F)
    {pass[feat] = NULL
    }else{
      fail[feat] = NULL
    }
    
    # Combined verdict:
    # If Shapiro PASSES -> normal
    # If Shapiro FAILS but practical thresholds pass -> practically normal
    # If both fail -> not normal
    verdict <- ifelse(
      sh$p.value > 0.05, "Normal",
      ifelse(practical, "Practically normal", "Not normal")
    )
    
    results <- rbind(results, data.frame(
      feature = feat,
      shapiro_p = round(sh$p.value, 4),
      skewness = round(sk, 3),
      kurtosis = round(kt, 3),
      practical = practical,
      final_verdict = verdict,
      stringsAsFactors = FALSE
    ))
  }
  return(list(results = results, fail = fail, pass = pass))
}
```


```{r}
output <- check_normality(class_0, continuous)
fail_class_0 <- output$fail
pass_class_0 <- output$pass
output$

output <- check_normality(class_1, continuous)
fail_class_1 <- output$fail
pass_class_1 <- output$pass

setdiff(names(pass_class_0), names(pass_class_1))
```
Features that passed normality check in class 1 also passed in class 0

```{r}
passed <- rbind(pass_class_0, pass_class_1)
failed <- rbind(fail_class_0, fail_class_1)
```

With that, lets prepare data for model. We take all normaly distributed continuous data and all discrete.

```{r}
train_data_nb <- train_data[, union(names(passed), discrete)]
test_data_nb <- test_data[, union(names(passed), discrete)]
```


#### Passed
Lets observe features that passed
```{r}
qqplot_df <- function(df){
  par(mfrow = c(1, 1))
  predictors <- names(df)
  for (feat in predictors) {
    qqnorm(df[[feat]], main = feat, cex.main = 0.8)
    qqline(df[[feat]], col = "red")
  }
}
```


```{r}
qqplot_df(passed)
```
#### Failed


```{r}
qqplot_df(failed)
```

## Model fit

```{r}
nb_model <- naiveBayes(
  as.factor(target) ~ .,
  data = train_data_nb,
  laplace = 1 # prevents a single rare value from dominating prediction
)
```

## Summary

### Metrics
```{r}

# Predict on test set
nb_pred <- predict(nb_model, newdata = test_data_nb, type = "class")
nb_prob <- predict(nb_model, newdata = test_data_nb, type = "raw")


tab_test <- table(nb_pred, as.factor(test_data_nb$target))
print(tab)

acc <- accuracy(tab_test)
prec <- precision(tab_test)

cat("Test Accuracy:", round(acc, 4), "\n")
cat("Test Precision:", round(prec, 4), "\n")


# Predict on training data
nb_train_pred <- predict(nb_model, newdata = train_data_nb, type = "class")

# Accuracy
tab_train <- table(nb_train_pred, as.factor(train_data_nb$target))
print(tab_train)

acc <- accuracy(tab_train)
prec <- precision(tab_train)
cat("Train Accuracy:", round(acc, 4), "\n")
cat("Train Precision:", round(prec, 4), "\n")
```

### Graphs
Lets take a look at the predictors and how they separate the classes

```{r}
for (fet in names(train_data_nb)){
  p <- ggplot(train_data_nb,
         aes(x = .data[[fet]], fill = as.factor(target))) +
    geom_density(alpha = 0.5) +
    labs(fill = "Target") +
    theme_minimal()
  print(p)
}
```


```{r}
library(pROC)

probs <- predict(nb_model, test_data_nb, type = "raw")[, "1"]
roc_obj <- roc(test_data_nb$target, probs)
plot(roc_obj, main = paste("AUC:", round(auc(roc_obj), 3)))
```



# LDA

## Assumptions of LDA
1. Binary outcome (already satisfied)
2. Normality within each class
3. Equal covariance matrices across classes (homogeneity)
4. No multicollinearity
5. No missing data
6. Sufficient sample size

## Preprocessing
LDA only work with continuous data which means we are splitting continuous again as we did in NB.

### Continuous/Discrete Split
For Naive Bayes we tryed to select continuous data by hand. Here we will try algorithmic approach for every average unique count >= 5 we define the features as discrete.


```{r}
predictors <- setdiff(names(train_data), "target")

continuous_alg <- c()
discrete_alg <- c()

for (feat in predictors) {
  n_unique <- length(unique(train_data[[feat]]))
  avg_count <- nrow(train_data) / n_unique
  
  if (avg_count >= 5) {
    discrete_alg <- c(discrete_alg, feat)
  } else {
    continuous_alg <- c(continuous_alg, feat)
  }
}
```

Lets compare this approach and our visual only


```{r}
cat("Manual approach: ", length(continuous), "\n")
cat("Alg approach: ", length(continuous_alg), "\n")

temp <- c()
for (feat in discrete) {
  temp <- c(temp, length(unique(train_data[[feat]])))
}
max(temp)

temp <- c()
for (feat in discrete_alg) {
  temp <- c(temp, length(unique(train_data[[feat]])))
}
max(temp)
```

We made the cut off for discrete feature having < 20 unique values, whereas with the algorithmic approach the cut off is < 185 🫢


```{r}
train_lda <- train_data[, c("target", continuous_alg)]
test_lda <- test_data[, c("target", continuous_alg)]
```



### Normality check
Same as with NB we need to check normality for the continuous features for each target class

```{r}
class_0 <- train_lda[train_lda$target == 0, ]
class_1 <- train_lda[train_lda$target == 1, ]

output <- check_normality(class_0, continuous_alg)
fail_class_0 <- output$fail
pass_class_0 <- output$pass

output <- check_normality(class_1, continuous_alg)
fail_class_1 <- output$fail
pass_class_1 <- output$pass


setdiff(names(pass_class_0), names(pass_class_1))
```

features that passed normality check in class 1 also passed in class 0

```{r}
passed <- rbind(pass_class_0, pass_class_1)
failed <- rbind(fail_class_0, fail_class_1)

length(passed)

train_lda <- train_lda[, c("target", names(passed))]
test_lda <- test_lda[, c("target", names(passed))]
```

Compare to NB where we had 16 after continuous and normality check here we have 13 features.


### Colinearity
Same as with logistic regression we will use VIF as metric if our data are colinear or not

Before doing VIF check

```{r}
length(names(train_lda))
```

After doing VIF check

```{r}
new_pred <- remove_vif(train_lda)
length(new_pred)
```

```{r}
train_lda <- train_lda[, c("target", new_pred)]
test_lda <- test_lda[, c("target", new_pred)]
```

### Homoscedasticity
First we can use Box's M test


```{r}
library(MASS)
library(biotools)

predictors <- setdiff(names(train_lda), "target")

boxM_result <- boxM(
  train_lda[, predictors],
  train_lda$target
)
print(boxM_result)
```
Box's M test rejected null hypothesis. We will still check the variance manually using boxplots and comparing variance of each features across classes

```{r}
predictors <- setdiff(names(train_lda), "target")

class_0 <- train_lda[train_lda$target == 0, ]
class_1 <- train_lda[train_lda$target == 1, ]

# Compare variance of each feature across classes
var_comparison <- data.frame(
  feature = predictors,
  var_class0 = sapply(predictors, function(f) var(class_0[[f]], na.rm = TRUE)),
  var_class1 = sapply(predictors, function(f) var(class_1[[f]], na.rm = TRUE))
)

var_comparison$ratio <- var_comparison$var_class1 / var_comparison$var_class0
# Ratio close to 1 = equal variance
# Ratio > 2 or < 0.5 = problematic

var_comparison <- var_comparison[order(abs(var_comparison$ratio - 1), decreasing = TRUE), ]
print(var_comparison, row.names = FALSE)

cat("\nProblematic (ratio > 2 or < 0.5):",
    sum(var_comparison$ratio > 2 | var_comparison$ratio < 0.5), "\n")

```
We can see no problematic features, lets also check boxplots


```{r}

par(mfrow = c(1, 2))

for (feat in predictors) {
  boxplot(
    train_lda[[feat]] ~ train_lda$target,
    main = feat, cex.main = 0.7,
    names = c("Loss", "Win"),
    col = c("lightblue", "lightcoral")
  )
}
```

Box's M test rejected the null hypothesis of equal covariance matrices (p < 0.001), however this test is known to be overly sensitive.Visual inspection and variance ratio analysis showed all features had variance ratios between 0.82 and 1.24, indicating the assumption is practically met.


## Model fit

```{r}
lda_model <- lda(
  as.factor(target) ~ .,
  data = train_lda
)
```


## Summary

```{r}
lda_train_pred <- predict(lda_model, train_clean)$class
train_tab <- table(lda_train_pred, as.factor(train_clean$target))
print(train_tab)

acc <- accuracy(train_tab)
prec <- precision(train_tab)

cat("Test Accuracy:", round(acc, 4), "\n")
cat("Test Precision:", round(prec, 4), "\n")

# Test accuracy
lda_test_pred <- predict(lda_model, test_clean)$class
test_tab <- table(lda_test_pred, as.factor(test_clean$target))
print(test_tab)

acc <- accuracy(test_tab)
prec <- precision(test_tab)

cat("Test Accuracy:", round(acc, 4), "\n")
cat("Test Precision:", round(prec, 4), "\n")

```


```{r}
library(MASS)


# Get projected values
lda_proj <- predict(lda_model, test_lda)

# Plot the projection
ggplot(data.frame(LD1 = lda_proj$x[,1], target = as.factor(test_lda$target)),
       aes(x = LD1, fill = target)) +
  geom_density(alpha = 0.5) +
  labs(title = "LDA Projection", x = "LD1", fill = "Target") +
  theme_minimal()
```

```{r}
coefs <- data.frame(
  feature = names(lda_model$scaling[,1]),
  weight = lda_model$scaling[,1]
)

ggplot(coefs, aes(x = reorder(feature, abs(weight)), y = weight)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "LDA Coefficients", x = "", y = "Weight") +
  theme_minimal()
```





