#Scenario 2
Parametric vs non-parametric models


Loading the libraries
```{r}
library(tidyverse)
```

Loading the preprocessed dataset
```{r}
df <- read.csv("processed.csv", sep=";", dec=",")
#removing the first column (exported indexing)
df <- df[2:ncol(df)]
```

```{r}

```

##Statistical assumptions
Since in this project, the main goal is binary classification (weather a team loses/wins), we are going to be using binary clasification models. For these models, there are a few assumptions we must check before proceeding with a certain model. These assumptions are:
- independence (each observation is independent - each of the observations is a sepparate new match)
- normality
- colinearity
- outliers
- homoscedasticity
- outliers (we have already checked in EDA, however, we might need to get rid of them based on the model we will use)
- feature scaling (not really a check, more of a data preparation)
- sample size (our sample size is 1216 obs.)
- linear relationship


Checking normality
```{r}
results <- list()
for(i in colnames(df)){
  x <- df[[i]]
  if(is.numeric(x)){
    results[i] = shapiro.test(x)$p.value
  }
}

bad_cols <- names(results)[results < 0.05]
length(bad_cols)


for(i in colnames(df)){
  x <- df[[i]]
  if(is.numeric(x)){
    qqnorm(x, main=i)
    qqline(x, col="red")
  }
  
}



```





