---
title: "Parametric_Models"
author: "Milan Mikuláš, Marek Vatrt"
date: "`r Sys.Date()`"
output: html_document
---

# Goal
In this section of our project, our goal is to build 3 non-parametric models, analyze the results and compare them with each other. To achieve the best results, we will firstly use default settings for creating the models and then apply hyperparameter tuning to get the best parameters (parameters with which the model has the highest accuracy on testing data).

# Loading libraries and data
```{r}
library(tidyverse)
library(tidymodels)

#vif values
library(car)

#Decision tree libs
library(rpart)
library(rpart.plot)
library(caret)

#random forest
library(randomForest)

#k-nn
library(class)

#metrics lib
library(Metrics)
#https://cran.r-project.org/web/packages/Metrics/Metrics.pdf

#clear env variables
rm(list = ls())
```


```{r}
df <- read.csv("processed.csv", sep=";", dec=",")
#removing the first column (exported indexing)
df <- df[2:ncol(df)]
df <- df %>%
  select(-early_advantage)
```


#Decision tree
First non-parametric model we chose is the decision tree. This choice was made because of the tree's simplicity and interpretability (although our dataset contains ~55 features, so the interpretability might be questionable, depending on how the tree will be created). Since decision trees can be used for regression as well as classification tasks, we can use it for our task of classifying win/loss of a team.

## Statistical assumptions
Statistical assumptions are taken from the 5th module.
### Independence
This dataset consists of random matches from the game League of Legends. Each of the games is independent of eachother, this is because each game is new. There are no rematch mechanics.
### Multicolinearity
Decision trees can handle multicolinearity, so we do not have to check for this (for random forest, we will)
### Outliers
Robust to outliers
### Sample size
May overfit on small datasets (that is what we want to test)
### Normality, Feature scaling, Homoscedasticity, Linear relationship
Not required

###Conclusion to statistical assumptions
Our dataset satysfies all the statistical assumptions that decision trees require. We can now move on to the implementation of the decision tree.

## Implementation
Setting up variables
```{r}
train_margin=0.7
```

Preparing the data
```{r}
#after running the model multiple times, we have decided that, for the decision trees, diff_inhibitor_takedowns and diff_baron_kills are very strong variables that made the trees stop almost instantly (depth=3)
#for the sake of exploring more relationships between features, we have decided to remove them and try again
df_dt <- df #%>%
  #select(-c(diff_inhibitor_takedowns, diff_baron_kills))

set.seed(69)
train_index <- sample(1:nrow(df_dt), train_margin * nrow(df_dt))

train_data <- df_dt[train_index, ]
test_data  <- df_dt[-train_index, ]
```

Sources:
https://www.rdocumentation.org/packages/rpart/versions/4.1.24/topics/rpart
https://www.rdocumentation.org/packages/rpart.plot/versions/3.1.3/topics/rpart.plot
https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict

For the implementation of the decision tree, we will be using the function rpart(). In this section, we will simply use the default parameters of rpart. We want to see how the default model compares to the hyperparameter-tuned model. Also, since decision trees can use Gini impurity and Entropy scores as the splitting criteria, we will also compare the accuracy of trees built with Gini and Entropy.

```{r}
#using lib rpart, we create trees for testing the gini inpurity and entropy
decision_tree_model_gini <- rpart(
  target ~ ., 
  data = train_data,
  #method of classfication
  method = "class",
  #parms is a list of set parameters, even if we just use 1 parameter, rpart expects a list()
  parms=list(split="gini"),
)

decision_tree_model_entropy <- rpart(
  target ~ ., 
  data = train_data,
  method = "class",
  parms=list(split="information"),
)

#plotting the trees we created
rpart.plot(decision_tree_model_gini,
           main="Gini Inpurity")
rpart.plot(decision_tree_model_entropy,
           main="Entropy")

#using predict() function, we use our test data to make predicitons
predictions_gini <- predict(decision_tree_model_gini, test_data, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy, test_data, type = "class")

#accuracy scores on test dataset for gini and entropy
cat("Accuracy on testing dataset (Gini)",round(mean(predictions_gini == test_data$target) * 100, 2),"%\n")
cat("Accuracy on testing dataset (Entropy)",round(mean(predictions_entropy == test_data$target) * 100, 2),"%")

```

```{r}

#predictions on train data - overfitting check
predictions_gini <- predict(decision_tree_model_gini, train_data, type = "class")
predictions_entropy <- predict(decision_tree_model_entropy, train_data, type = "class")

#accuracy on training dataset for gini and entropy
cat("Accuracy on training dataset (Gini)",round(mean(predictions_gini == train_data$target) * 100, 2),"%\n")
cat("Accuracy on training dataset (Entropy)",round(mean(predictions_entropy == train_data$target) * 100, 2),"%")
```
As we can see, the decision tree, without even tuning its hyperparameters, got scores of 95% for accuracy. The more accurate out of the Gini Inpurity and Entropy methods is the Gini Impurity. For that reason, we will be using Gini Impurity scoring for tuning the hyperparameters. 

Another fact to mention is that the decision tree is not overfitted. Actually, for gini impurity, the accuracy score falls behind on the training dataset by about 0.3%. On the contrary, Entropy score rises on the training dataset. Both of these scores, however, are very similar to those of the testing dataset scores. From this we can conclude that the decision trees are not overfitted.

Since we have decided on the scoring metric we will be using, we can now move on to tuning the decision tree's hyperparameters. The rpart() function for creating the decision trees does not have built in hyperparameters unfortunately. However, we can use rpart.control() to modify the hyperparameters. The creation and training of decision trees is not as computanionaly expensive as creating a random forest and tuning it's hyperparameters, so for the DT, we have decided to use grid search to search for the best hyperparameters.

The probability of finding a better decision tree is quite slim however. This is due to the fact that the basic rpart() function has already found a tree that has a depth of only 3.

## Hyperparameter tuning
```{r}
set.seed(69)

train_data$target <- as.factor(train_data$target)
test_data$target <- as.factor(test_data$target)

k_folds <- 5

folds <- createFolds(
  train_data$target,
  k = k_folds,
  list = TRUE,
  returnTrain = FALSE
)

dt_grid <- expand.grid(
  cp = c(0.0001, 0.001, 0.005, 0.01, 0.02, 0.05),
  maxdepth = c(3, 5, 7, 10, 15, 20, 30),
  minsplit = c(5, 10, 20, 30, 40),
  minbucket = c(2, 5, 10, 15)
)

best_accuracy <- 0
best_params <- NULL
results <- data.frame()

for (i in 1:nrow(dt_grid)) {
  
  fold_accuracies <- numeric(k_folds)
  
  for (fold_id in 1:k_folds) {
    
    validation_index <- folds[[fold_id]]
    
    cv_train_data <- train_data[-validation_index, ]
    cv_validation_data <- train_data[validation_index, ]
    
    dt_model <- rpart(
      target ~ .,
      data = cv_train_data,
      method = "class",
      parms = list(split = "gini"),
      control = rpart.control(
        cp = dt_grid$cp[i],
        maxdepth = dt_grid$maxdepth[i],
        minsplit = dt_grid$minsplit[i],
        minbucket = dt_grid$minbucket[i]
      )
    )
    
    predictions <- predict(
      dt_model,
      newdata = cv_validation_data,
      type = "class"
    )
    
    fold_accuracies[fold_id] <- mean(predictions == cv_validation_data$target)
  }
  
  mean_cv_accuracy <- mean(fold_accuracies)
  
  results <- rbind(
    results,
    data.frame(
      cp = dt_grid$cp[i],
      maxdepth = dt_grid$maxdepth[i],
      minsplit = dt_grid$minsplit[i],
      minbucket = dt_grid$minbucket[i],
      cv_accuracy = mean_cv_accuracy
    )
  )
  
  if (mean_cv_accuracy > best_accuracy) {
    best_accuracy <- mean_cv_accuracy
    best_params <- dt_grid[i, ]
  }
}

results <- results[order(results$cv_accuracy, decreasing = TRUE), ]

best_dt_model <- rpart(
  target ~ .,
  data = train_data,
  method = "class",
  parms = list(split = "gini"),
  control = rpart.control(
    cp = best_params$cp,
    maxdepth = best_params$maxdepth,
    minsplit = best_params$minsplit,
    minbucket = best_params$minbucket
  )
)

test_predictions <- predict(
  best_dt_model,
  newdata = test_data,
  type = "class"
)

train_predictions <- predict(
  best_dt_model,
  newdata = train_data,
  type = "class"
)

test_accuracy <- mean(test_predictions == test_data$target)
train_accuracy <- mean(train_predictions == train_data$target)

print(results)
print(best_params)

cat("Best CV accuracy:", round(best_accuracy * 100, 2), "%\n")
cat("Best cp:", best_params$cp, "\n")
cat("Best maxdepth:", best_params$maxdepth, "\n")
cat("Best minsplit:", best_params$minsplit, "\n")
cat("Best minbucket:", best_params$minbucket, "\n")
cat("Training accuracy:", round(train_accuracy * 100, 2), "%\n")
cat("Testing accuracy:", round(test_accuracy * 100, 2), "%\n")
```
```{r}
rpart.plot(best_dt_model,
           main="Gini Inpurity - best hyperparameters model")
rpart.plot(decision_tree_model_gini,
           main="Gini Inpurity - default model")
```
As we have suspected, the best model, even with tweaking the hyperparameters is still the one created using default rpart() configuration. The model with tuned hyperparameters achieved better performance on the training dataset, however, the accuracy score for the testing dataset remains the same.


# Random forest
Building upon the tree models, due to the unexpectedly high accuracy score of decision tree from both the test and train datasets, we have decided to choose random forest.

## Statistical assumptions
Statistical assumptions are taken from the 5th module.
### Independence
This dataset consists of random matches from the game League of Legends. Each of the games is independent of eachother, this is because each game is new. There are no rematch mechanics.
### Multicolinearity
Although random forest can handle multicolinearity, it is stated that colinearity could affect feature importance. 
Due to this reason, we have decided to get rid of features that have too big of multicolinarity between each other.

Since we have already cut the features with above 0.8 value on the correlation matrix during EDA and preprocessing, this time we will focus on the individual VIF values of the features. We will continue to remove the feature with the biggest VIF value until the maximum VIF value is below 10.

Sources:
https://www.rdocumentation.org/packages/regclass/versions/1.6/topics/VIF

```{r, warning=FALSE}
#copy of the current df, due to the elimination of variables
random_forest_df <- df

#upon consideration, we have decided to 
#random_forest_df <- random_forest_df %>%
#  select(-diff_inhibitor_takedowns)

#Since our task is to classify data, we will be fitting a logistic regression model into the VIF function
vif_model <- glm(target ~ ., data = random_forest_df, family = "binomial")
summary(vif_model)
vif_values <- vif(vif_model)
#vif_values
names(vif_values)[which.max(vif_values)]
max(vif_values)

features_removed <- 0

while (max(vif_values)>=10) {
  features_removed <- features_removed + 1
  feature_to_eliminate <- names(vif_values)[which.max(vif_values)]
  random_forest_df <- random_forest_df %>%
    select(-feature_to_eliminate)
  vif_model <- glm(target ~ ., data = random_forest_df, family = "binomial")
  vif_values <- vif(vif_model)
  print(names(vif_values)[which.max(vif_values)])
  print(max(vif_values))
}

cat("Number of features removed:",features_removed)
vif_values

```
We ended up only removing 2 features, however, the first feature we removed had a VIF score of over 4 billion, which is an astronomically high multicolinearity score. The second feature was over 100 million. From the scores alone, we can asses that it was a good decision to remove them from the dataset before creating a random forest model.

### Outliers
Robust to outliers
### Sample size
Performs better with a larger dataset. Our dataset is not very large (~1200 rows), however, due to the accuracy of the decision tree we tested, we believe that the random forest will still perform well.
### Normality, Feature scaling, Homoscedasticity, Linear relationship
Not required

## Implementation
Setting up variables
```{r}
train_margin=0.7
```


```{r}
random_forest_df$target <- as.factor(random_forest_df$target)

set.seed(69)
train_index <- sample(1:nrow(random_forest_df), train_margin * nrow(random_forest_df))

train_data <- random_forest_df[train_index, ]
test_data  <- random_forest_df[-train_index, ]

# Build Random Forest model
random_forest_model <- randomForest(
  target ~ .,
  data = train_data,
  replace = TRUE,
  importance = TRUE
)

# Show model summary
random_forest_model

# Predict on test data
rf_predictions <- predict(
  random_forest_model,
  newdata = test_data,
  type = "class"
)

# Confusion matrix
table(
  Predicted = rf_predictions,
  Actual = test_data$target
)

# Accuracy
rf_accuracy <- mean(rf_predictions == test_data$target)

cat("Random Forest accuracy:", round(rf_accuracy * 100, 2), "%")
```
As we can see. Just by using the default values (+ setting the replacement and importance to TRUE), we got up to 95.34% accuracy. This score is outstanding, but we are wondering if we can do even better by searching through multiple combinations of hyperparameters to find a better result. We have decided to use random search for hyperparameters. We could also use grid search, however, finding the best combinations of hyperparameters with grid search could become very exhaustive, especially with the ~55 features we have. Random search could find a better combination of hyperparameters much faster because it does not waste time trying every combination that exists.

At first, we tried using the train() function, however, the only hyperparameter we are able to tweak is the **mtry** parameter. Mtry is important, but this kind of random testing only provided minor improvements (around 0.5% accuracy). We are wondering if it is possible to go beyond that. In this instance, we will not be using a built-in function, instead we will try to do our own version of a randomized search, where we try to tweak all the numeric parameters of the random forest - mtree, mtry, nodesize, maxnodes using the randomForest library and cross validation. 
```{r}
set.seed(69)

#n_features <- ncol(train_data) - 1
rf_iterations <- 200

#we will create a "table" for each iteration of the random forest model
random_grid <- data.frame(
  #selecting the number of features, from 1 to 20, we used to include all features, but the accuracy score actually decreased
  mtry = sample(1:30, rf_iterations, replace = TRUE),
  
  #selecting the number of trees, at first we wanted to choose 100:1000, but then decided to go by sequence of 25 trees from 100 to 1000 trees
  # we thought that 100:1000 could 
  ntree = sample(seq(100, 1000, by = 50), rf_iterations, replace = TRUE),
  
  #for nodesize, we decided to not start from 1, because that could cause the trees to overfit quite easily if the nodes only needed 1 or 2 samples to split
  nodesize = sample(20:100, rf_iterations, replace = TRUE),
  
  #for max nodes, we decided for a sequence of 10 to 100 by 5, 100 might be a bit too much and will overfit the data, we will see
  maxnodes = sample(seq(10,40, by=2), rf_iterations, replace = TRUE)
)

#if any of the hyperparameter sampling ranges changed (for example maxnodes from 10-100 to 10-50), it is because we found that the randomization caused overfitting or we found a better range for the hyperparameter 
#we had to reduce the maxnodes, because the tree started overfitting at the bigger number of nodes, also we changed nodesize from 5:20, to 20:100, which helped with overfitting
#we managed to reduce overfitting this way by 1.6% (99.8 -> 98.24)
```

```{r}
set.seed(69)

train_data$target <- as.factor(train_data$target)
test_data$target <- as.factor(test_data$target)

best_accuracy <- 0
best_rf_model <- NULL
best_params <- NULL
results <- data.frame()

k_folds <- 5

folds <- createFolds(
  train_data$target,
  k = k_folds,
  list = TRUE,
  returnTrain = FALSE
)

for (i in 1:nrow(random_grid)) {
  
  cv_accuracies <- c()
  
  for (fold_id in 1:k_folds) {
    
    validation_index <- folds[[fold_id]]
    
    cv_train_data <- train_data[-validation_index, ]
    cv_validation_data <- train_data[validation_index, ]
    
    cv_train_data$target <- as.factor(cv_train_data$target)
    cv_validation_data$target <- as.factor(cv_validation_data$target)
    
    rf_model <- randomForest(
      target ~ .,
      data = cv_train_data,
      mtry = random_grid$mtry[i],
      ntree = random_grid$ntree[i],
      nodesize = random_grid$nodesize[i],
      maxnodes = random_grid$maxnodes[i],
      replace = TRUE,
      importance = TRUE
    )
    
    pred <- predict(rf_model, newdata = cv_validation_data)
    accuracy <- mean(pred == cv_validation_data$target)
    
    cv_accuracies <- c(cv_accuracies, accuracy)
  }
  
  mean_cv_accuracy <- mean(cv_accuracies)
  
  results <- rbind(
    results,
    data.frame(
      mtry = random_grid$mtry[i],
      ntree = random_grid$ntree[i],
      nodesize = random_grid$nodesize[i],
      maxnodes = random_grid$maxnodes[i],
      cv_accuracy = mean_cv_accuracy
    )
  )
  
  if (mean_cv_accuracy > best_accuracy) {
    best_accuracy <- mean_cv_accuracy
    best_params <- random_grid[i, ]
  }
}

results <- results[order(results$cv_accuracy, decreasing = TRUE), ]
results
best_params
best_accuracy
```

```{r}
best_rf_model <- randomForest(
  target ~ .,
  data = train_data,
  mtry = best_params$mtry,
  ntree = best_params$ntree,
  nodesize = best_params$nodesize,
  maxnodes = best_params$maxnodes,
  replace = TRUE,
  importance = TRUE
)
```

```{r}
test_pred <- predict(best_rf_model, newdata = test_data)
test_accuracy <- mean(test_pred == test_data$target)

train_pred <- predict(best_rf_model, newdata = train_data)
train_accuracy <- mean(train_pred == train_data$target)

cat("Best CV accuracy:", round(best_accuracy * 100, 2), "%\n")
cat("Training accuracy:", round(train_accuracy * 100, 2), "%\n")
cat("Testing accuracy:", round(test_accuracy * 100, 2), "%\n")
```

The best outcome we got from this hyperparameter tuning after we tried changing all the randomised ranges of the hyperparameters was 96.99%. That is an upgrade of 1.1% compared to the train() function. What we have not done yet is checking for overfitting. We will now test the best model on both the train dataset and the test dataset and compare their accuracy.


## Hyperparameter tuning conclusion
The most succesful model came from our own implementation of random search with an accuracy of 96.99%. Even though it might not seem like a substanstial difference, the "baseline" for this model was already very high at 95.34%. 

### Feature importance
Now that we have found the best model via hyperparameter tuning, we will look into what are the most important, or valuable features for splitting the trees.

Sources:
https://www.rdocumentation.org/packages/randomForest/versions/4.7-1.2/topics/varImpPlot

```{r, fig.width=16, fig.height=8}
varImpPlot(
  best_rf_model,
  n.var = 10,
  main = "Random Forest Feature Importance"
)
```
As we can see, the same feature that dominated the decision tree dominated again in the random forest. 

diff_inhibitor_takedowns, without being a data leakage, is the most significant and influential variable for predicting the outcome of the match. It has the highest MeanDecreaseAccuracy, meaning that randomly permuting this feature causes the largest drop in classification accuracy. It also has the highest MeanDecreseGini, which means that it contributes the most to the reduction of gini impurity at a split.

We debated wether to remove it, simply because it carries a lot of influencer over the outcome. However, as the decision tree has proved, even if we were to remove it, other features can still classify the outcome of the matches with a great accuracy. In the end, we have decided to keep it, because it is a part of the dataset and it is not considered a data leakage. 



#k-NN
As the final non-parametric model we have chosen k-NN. At first we wanted to use XGBoost, because we thought that the decision tree and random forest would not be as powerful, so we wanted to use boosting to combine the "weaker" trees into a better one. However, the accuracy of the tree models turned out to be excellent, that is why we decided to move on from tree algorithms.

k-NN is simple, and unlike tree algorithms, that split the feature space based on conditions, k-NN uses the similarity of data to make decisions. We wanted to choose it due to the use of a different technique. However, k-NN has a defining problem, sometimes called the "curse of dimensionality". What this means is that if a dataset contains too many features, the distances between individual features can become similar, which can cause the model to predict unreliably. This is a big problem for our dataset, which contains ~55 features.

To counter this problem, we are planning on implementing a simple feature selection, to select the best 5-15 features. We plan to do this using the most random forest's feature importance. Using the top 15 features, we will then observe how the accuracy of k-NN increases/decreases when we try using 14,13,12,..,5 features.

What we found from using the tree models is the fact that our dataset contains a small set of features, which carry the most influence over the outcome of the prediction. We want to see exactly how much the number of features changes the accuracy and if a different algorithm can bring even more accuracy to our predictions.

## Statistical assumptions
Sources for statistical assumptions of k-NN: \n
https://medium.com/analytics-vidhya/k-nearest-neighbors-algorithm-7952234c69a4 \n

### Independence
This dataset consists of random matches from the game League of Legends. Each of the games is independent of eachother, this is because each game is new. There are no rematch mechanics.
### Multicolinearity

### Outliers
Sensitive to outliers.
```{r}
count_outliers_iqr <- function(df) {
  numeric_cols <- sapply(df, is.numeric)
  df_num <- df[, numeric_cols, drop = FALSE]
  
  result <- data.frame(
    column = character(),
    number_of_outliers = integer()
  )
  
  for (col in names(df_num)) {
    Q1 <- quantile(df_num[[col]], 0.25, na.rm = TRUE)
    Q3 <- quantile(df_num[[col]], 0.75, na.rm = TRUE)
    IQR_value <- Q3 - Q1
    
    lower <- Q1 - 1.5 * IQR_value
    upper <- Q3 + 1.5 * IQR_value
    
    count <- sum(df_num[[col]] < lower | df_num[[col]] > upper, na.rm = TRUE)
    
    result <- rbind(result, data.frame(
      column = col,
      number_of_outliers = count
    ))
  }
  
  result <- result[order(-result$number_of_outliers), ]
  
  print(result)
}
```

```{r}
count_outliers_iqr(df)       # rows with at least one outlier
```


### Sample size
May overfit on small datasets (that is what we want to test)
### Normality, Homoscedasticity, Linear relationship
Not required
### Feature scaling
Due to k-NN being a distance-based method of classification

```{r}
#k-nn
library(class)
library(kknn)
```


##Implementation

###Testing with all features
```{r}
knn_df <- df
knn_df$target <- as.factor(knn_df$target)

set.seed(69)
train_index <- sample(1:nrow(knn_df), train_margin * nrow(knn_df))

knn_train_data <- knn_df[train_index, ]
knn_test_data  <- knn_df[-train_index, ]
```

Base model with default hyperparameters
```{r}
library(kknn)
model <- kknn(
          target ~ .,
          knn_train_data,
          knn_test_data
        )

pred <- predict(model)
acc <- mean(pred == knn_test_data$target)
acc
```

### Outliers

### Multicollinearity
Multicollinearity can affect k-NN performance, we can try eliminating the features with high multicollinearity, just like we did in RF. 

###Hyperparameter tuning
The hyperparameters we will be tuning to find the best k-NN model is the k value, meaning the number of neighbours needed for classifying a point, distance, and kernel. We can choose distances 1,2,3 - Manhattan, Euclidian and Minkowski. There are 10 kernels available for kknn, we chose to try all of them. The final number of combinations adds up to 1500.
```{r}
#we would like to avoid using k that is too big, since it could easily misidentify values, however, we want to test it
#upon testing it on out implementation and using train() function, we have decided cut out hyperparameters to only the best ranges
k_values <- 1:35
distance_metrics <- 1:2
kernels <- c(
  "rectangular",
  "epanechnikov",
  "biweight",
  "rank"
)

k_folds <- 5

set.seed(69)
folds <- createFolds(
  knn_train_data$target,
  k = k_folds,
  list = TRUE,
  returnTrain = FALSE
)

best_k <- 0
best_metric <- 0
best_kernel <- NULL
knn_best_model <- NULL
best_accuracy <- 0
results <- data.frame()
```

We will iterate through all the combinations - grid search, since there is "only" 1500 combinations, doing this will not be as computationaly expensive as random forest, for which we opted to use random search.
###NEW HYPERSESRCH
```{r}
for (k in k_values) {
  for (kernel in kernels) {
    for (metric in distance_metrics) {
      
      fold_accuracies <- numeric(k_folds)
      
      for (fold_id in 1:k_folds) {
        
        validation_index <- folds[[fold_id]]
        
        cv_train_data <- knn_train_data[-validation_index, ]
        cv_validation_data <- knn_train_data[validation_index, ]
        
        model <- kknn(
          target ~ .,
          train = cv_train_data,
          test = cv_validation_data,
          k = k,
          distance = metric,
          kernel = kernel,
          scale = TRUE
        )
        
        predictions <- predict(model)
        fold_accuracies[fold_id] <- mean(predictions == cv_validation_data$target)
      }
      
      cv_accuracy <- mean(fold_accuracies)
      
      results <- rbind(
        results,
        data.frame(
          k = k,
          kernel = kernel,
          distance = metric,
          cv_accuracy = cv_accuracy
        )
      )
      
      if (cv_accuracy > best_accuracy) {
        best_accuracy <- cv_accuracy
        best_k <- k
        best_metric <- metric
        best_kernel <- kernel
      }
    }
  }
}

results <- results[order(results$cv_accuracy, decreasing = TRUE), ]

results
```

```{r}
knn_best_model <- kknn(
  target ~ .,
  train = knn_train_data,
  test = knn_test_data,
  k = best_k,
  distance = best_metric,
  kernel = best_kernel,
  scale = TRUE
)

final_predictions <- predict(knn_best_model)
final_accuracy <- mean(final_predictions == knn_test_data$target)

cat("Best CV accuracy:", round(best_accuracy * 100, 2), "%\n")
cat("Best k:", best_k, "\n")
cat("Best distance metric:", best_metric, "\n")
cat("Best kernel:", best_kernel, "\n")
cat("Final test accuracy:", round(final_accuracy * 100, 2), "%\n")
```


```{r}
# library(caret)
# library(kknn)
# 
# set.seed(123)
# 
# control <- trainControl(
#   method = "cv",
#   number = 5
# )
# 
# knn_grid <- expand.grid(
#   kmax = 1:50,
#   distance = 1:3,
#   kernel = c(
#     "rectangular",
#     "triangular",
#     "epanechnikov",
#     "biweight",
#     "triweight",
#     "cos",
#     "inv",
#     "gaussian",
#     "rank",
#     "optimal"
#   )
# )
# 
# knn_tuned_model <- train(
#   target ~ .,
#   data = knn_train_data,
#   method = "kknn",
#   trControl = control,
#   tuneGrid = knn_grid,
#   preProcess = c("center", "scale"),
#   metric = "Accuracy"
# )
# 
# best_params <- knn_tuned_model$bestTune
# 
# final_predictions <- predict(
#   knn_tuned_model,
#   newdata = knn_test_data
# )
# 
# final_accuracy <- mean(final_predictions == knn_test_data$target)
# 
# results <- knn_tuned_model$results
# results <- results[order(results$Accuracy, decreasing = TRUE), ]
# 
# print(results)
# 
# cat("Best CV accuracy:", round(max(knn_tuned_model$results$Accuracy) * 100, 2), "%\n")
# cat("Best k:", best_params$kmax, "\n")
# cat("Best distance metric:", best_params$distance, "\n")
# cat("Best kernel:", best_params$kernel, "\n")
# cat("Final test accuracy:", round(final_accuracy * 100, 2), "%\n")
```

We also tried to use train() and trainControl() to control the cross validation and training of the models. Due to the different creation of folds for the training and validation of the models, the outcomes slightly differed. We achieved similar results in terms of accuracy, however, the training time of all the models took 5x longer using the train() function (2 hours), so we have decided to choose our implementation.


## Feature selection
Since our dataset contains ~55 features, we have decided to try using a dimensionality reduction method to lower the number of features. For this, we have chosen PCA. 

```{r}
set.seed(69)

knn_train_data$target <- as.factor(knn_train_data$target)
knn_test_data$target <- as.factor(knn_test_data$target)

x_train <- knn_train_data[, names(knn_train_data) != "target"]
y_train <- knn_train_data$target

x_test <- knn_test_data[, names(knn_test_data) != "target"]
y_test <- knn_test_data$target

#accuracy was the highest for 0.4 threshold, therefore we went lower
#in the end, the accuracy stayed the biggest for 0.4, 
pca_thresholds <- c(0.2,0.25,0.3,0.35,0.4,0.45, 0.5,0.55)

k_values <- 15:45
distance_metrics <- 1:2
kernels <- c(
  "rectangular",
  "epanechnikov",
  "biweight",
  "inv",
  "rank"
)

k_folds <- 5

folds <- createFolds(
  y_train,
  k = k_folds,
  list = TRUE,
  returnTrain = FALSE
)

results <- data.frame()

best_cv_accuracy <- 0
best_threshold <- NULL
best_k <- NULL
best_distance <- NULL
best_kernel <- NULL
best_components <- NULL
```

```{r}
for (threshold in pca_thresholds) {
  
  for (k in k_values) {
    for (distance in distance_metrics) {
      for (kernel in kernels) {
        
        fold_accuracies <- numeric(k_folds)
        
        for (fold_id in 1:k_folds) {
          
          validation_index <- folds[[fold_id]]
          
          cv_x_train <- x_train[-validation_index, ]
          cv_y_train <- y_train[-validation_index]
          
          cv_x_validation <- x_train[validation_index, ]
          cv_y_validation <- y_train[validation_index]
          
          # PCA must be fitted only on CV training data
          pca_model <- preProcess(
            cv_x_train,
            method = c("center", "scale", "pca"),
            thresh = threshold
          )
          
          cv_x_train_pca <- predict(pca_model, cv_x_train)
          cv_x_validation_pca <- predict(pca_model, cv_x_validation)
          
          cv_train_pca <- data.frame(
            cv_x_train_pca,
            target = cv_y_train
          )
          
          cv_validation_pca <- data.frame(
            cv_x_validation_pca,
            target = cv_y_validation
          )
          
          model <- kknn(
            target ~ .,
            train = cv_train_pca,
            test = cv_validation_pca,
            k = k,
            distance = distance,
            kernel = kernel
          )
          
          predictions <- predict(model)
          
          fold_accuracies[fold_id] <- mean(
            predictions == cv_validation_pca$target
          )
        }
        
        mean_cv_accuracy <- mean(fold_accuracies)
        
        results <- rbind(
          results,
          data.frame(
            pca_threshold = threshold,
            k = k,
            distance = distance,
            kernel = kernel,
            cv_accuracy = mean_cv_accuracy
          )
        )
        
        if (mean_cv_accuracy > best_cv_accuracy) {
          best_cv_accuracy <- mean_cv_accuracy
          best_threshold <- threshold
          best_k <- k
          best_distance <- distance
          best_kernel <- kernel
        }
      }
    }
  }
}
```

```{r}
best_pca_model <- preProcess(
  x_train,
  method = c("center", "scale", "pca"),
  thresh = best_threshold
)

x_train_pca <- predict(best_pca_model, x_train)
x_test_pca <- predict(best_pca_model, x_test)

knn_train_pca <- data.frame(
  x_train_pca,
  target = y_train
)

knn_test_pca <- data.frame(
  x_test_pca,
  target = y_test
)

final_model <- kknn(
  target ~ .,
  train = knn_train_pca,
  test = knn_test_pca,
  k = best_k,
  distance = best_distance,
  kernel = best_kernel
)

test_predictions <- predict(final_model)
test_accuracy <- mean(test_predictions == knn_test_pca$target)

train_model <- kknn(
  target ~ .,
  train = knn_train_pca,
  test = knn_train_pca,
  k = best_k,
  distance = best_distance,
  kernel = best_kernel
)

train_predictions <- predict(train_model)
train_accuracy <- mean(train_predictions == knn_train_pca$target)

results <- results[order(results$cv_accuracy, decreasing = TRUE), ]

print(results)

cat("Best PCA threshold:", best_threshold, "\n")
cat("Number of PCA components:", ncol(x_train_pca), "\n")
cat("Best k:", best_k, "\n")
cat("Best distance:", best_distance, "\n")
cat("Best kernel:", best_kernel, "\n")
cat("Best CV accuracy:", round(best_cv_accuracy * 100, 2), "%\n")
cat("Training accuracy:", round(train_accuracy * 100, 2), "%\n")
cat("Testing accuracy:", round(test_accuracy * 100, 2), "%\n")
cat("Train-test gap:", round((train_accuracy - test_accuracy) * 100, 2), "%\n")
```









